Inference using context-dependent models of mutation

### 2017-06-04 writeups/context-dependence.tex
various

@@

      \item[(repair)] In addition to the above, a collection of motifs associated with known biochemical causes
          curated from the literature:
          deamination due to repair of UV-induced pyramidine dimers ($\mu_{UV}$: (\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
          an interaction between CpG and pyramidine dimers ($\mu_{mUV}$: \nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
          repair of guanine adducts at CpG sites ($\mu_g$: \nC\nG$\to$\nC\nT, \citet{tobacco});
          deamination by AID ($\mu_{AID}$: \nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
          deamination by members of the APOBEC family ($\mu_{APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
          and repair by error-prone polymerases pol $\eta$ ($\mu_\eta$: \nT\nA$\to$\nT\nG, \citet{pol_eta});
          and pol $\iota$ ($\mu_{\iota}$: \nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
          \begin{center}
            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
                \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
                \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
                \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA  &   \nT\nT\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
                \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT  &   \nT\nT\nT   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
                \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA  &   \nT\nG\nA   &  $\mu_{APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
                \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT  &   \nT\nG\nT   &  $\mu_{APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
                \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
                \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA  &   \nA\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
                \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
                \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA  &   \nA\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
                \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA    &    \nA\nG      &  $\mu_{\iota1}$ & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
                \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT    &    \nC\nT      &  $\mu_{\iota1}$ & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
                \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
                \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
                           &               &                 &           &                &                 & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
                           &               &                 &           &                &                 & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
            \end{tabular}
          \end{center}

@@

@@

>>>>

\newcommand{\APOBEC}{\text{\tiny APOBEC}}
\newcommand{\AID}{\text{\tiny AID}}

@@

      \item[(repair)] In addition to the above, a collection of motifs associated with known biochemical causes
          curated from the literature:
          deamination due to repair of UV-induced pyramidine dimers ($\mu_{UV}$: (\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
          an interaction between CpG and pyramidine dimers ($\mu_{mUV}$: \nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
          repair of guanine adducts at CpG sites ($\mu_g$: \nC\nG$\to$\nC\nT, \citet{tobacco});
          deamination by AID ($\mu_{AID}$: \nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
          deamination by members of the APOBEC family ($\mu_{\APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
          and repair by error-prone polymerases pol $\eta$ ($\mu_\eta$: \nT\nA$\to$\nT\nG, \citet{pol_eta});
          and pol $\iota$ ($\mu_{\iota}$: \nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):

          \noindent
      \begin{center}
          \begin{tabular}{c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c|c@{\;\;$\to$\;\;}c@{\;\; {\small at rate}\;\; }c}
            \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG       &   \nC\nT       &  $\mu_{g}$       & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG       &   \nA\nG       &  $\mu_{g}$       & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA    &   \nT\nT\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT    &   \nT\nT\nT    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA    &   \nT\nG\nA    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT    &   \nT\nG\nT    &  $\mu_{\APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA    &   \nA\nA\nA    &  $\mu_{\APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA    &   \nT\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA    &   \nA\nC\nA    &  $\mu_{\APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ \\
            \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA       &    \nA\nG      &  $\mu_{\iota1}$  & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{\AID}$ \\
            \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT       &    \nC\nT      &  $\mu_{\iota1}$  & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{\AID}$ \\
            \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ & \nT\nG\nC\nT &  \nT\nG\nT\nT  &  $\mu_{\AID}$    & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{\AID}$ \\
            \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ & \nT\nG\nC\nC &  \nT\nG\nT\nC  &  $\mu_{\AID}$    & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{\AID}$ 
        \end{tabular}
      \end{center}

@@

\begin{figure}
    \begin{center} 
        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors}
    \end{center} 
    \caption{
        Estimated average mutation rates
        for the motifs of Table \ref{tab:hominid_model}
        since the common ancestor of humans and chimpanzee.
        Rates are scaled so that the mean time to common ancestor is 0.001 units --
        if we take this to be 6 million years \citep{scally2012insights}
        then rates are in units of mutations per 6,000 years.
        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
        Figure \ref{sfig:biochem_results_zoom} shows the smaller rates in more detail.
        \label{fig:biochem_results}}
\end{figure}

@@

\begin{figure}
    \begin{center} 
        \includegraphics[width=\textwidth]{writeup-plots/biochem_mcmc_posteriors_zoomed}
    \end{center} 
    \caption{
        As in Figure \ref{sfig:biochem_results}, but with reduce vertical axis range
        to see in more detail:
        estimated average mutation rates
        for the motifs of Table \ref{tab:hominid_model}
        since the common ancestor of humans and chimpanzee.
        Rates are scaled so that the mean time to common ancestor is 0.001 units --
        if we take this to be 6 million years \citep{scally2012insights}
        then rates are in units of mutations per 6,000 years.
        Shown are posterior medians (points) and 95\% credible intervals (lines, which are quite small in all cases).
        \label{fig:biochem_results_zoom}}
\end{figure}
